### 2015-04-07 cug_paper/01_introduction.tex
Paper: Minor edits

Most HPC applications' I/O exhibit a very `bursty' pattern. That is, there will be long periods with no file output while the application is in a compute phase, followed by a period of intense output and no computation when the application writes its checkpoint or output files.  At ORNL, the filesystem is a center-wide resource shared with other systems including a Cray XC-30, analysis clusters, and WAN-connected data mover nodes. Each of these competing systems place a varying load on the filesystem such that Titan's users' write performance will vary dramatically within a job and between jobs.

>>>>

Most HPC applications' I/O exhibit a very `bursty' pattern. That is, there will be long periods with no file output while the application is in a compute phase, followed by a period of intense output and no computation when the application writes its checkpoint or output files.  At ORNL, the filesystem is a center-wide resource shared with multiple systems including a Cray XK7, a Cray XC30, analysis clusters, and WAN-connected data mover nodes. Each of these competing systems place a varying load on the filesystem and thus the performance that individual users see can vary dramatically within a job and between jobs.

### 2015-04-07 cug_paper/02_design.tex
Paper:  More edits & proofreading.  Last of the /todo boxes removed.
Paper: Minor edits

\subsection{Overview}
\label{subsec:overview}
To test this idea, we created a synthetic benchmark application that consists of two executables: the main application which uses MPI and simulates the I/O patterns of a typical HPC application and a daemon process that is started on each compute node.  The daemon has two tasks: manage the GPU memory and copy data from GPU memory out to a file.  Note that the daemon does \emph{not} copy the data into the GPU memory; that is handled by the individual ranks of the main application. There needs to be exactly one daemon process on each compute node regardless of how many MPI ranks are running on each node.\footnote{Due to the limitations of aprun, starting this daemon is somewhat involved.  See section \ref{subsec:cray-specific}.}

@@

The daemon maintains a pair of lists.  One list is the blocks of GPU memory that have been assigned to the ranks of the main application, and which the ranks are presumably copying data into.  The second list is for all the blocks of memory for which the `write complete' message (step e) has been received.  When the daemon thread that handles CCI messages receives a `write done' message for a block, it moves that block from the first list to the second list.  It then notifies the background write thread(s) that the block is now ready to be written to the filesystem. All daemon threads are pinned to unused integer cores.

@@

There are a few peculiarities of the Cray environment that need to be taken in to account in order for this software to work properly.  The first peculiarity has to do with limiting access to the GPU and associated hardware.  The CUDA runtime can be configured to limit access to the GPU to a single thread, to multiple threads from a single process, or to multiple threads from multiple processes.  On Titan, if no special options are given to qsub, the CUDA runtime will be configured for a single thread.  Since software described in this paper requires access to the GPU from multiple processes, users must switch to the appropriate compute mode.  This is done by passing the flag ``l feature=gpudefault'' to qsub.\footnote{It may seem strange to have to explicitly switch to a mode called `gpudefault'.  The naming scheme comes from the CUDA documentation.  On a standard CUDA install, the GPU would in fact default to allowing access from multiple processes.  Titan is configured differently, however.  Thus the need to explicitly tell qsub to switch the GPU's to the `default' mode.}

The second peculiarity has to do with actually starting the daemon process.  Normally, executables are started on the compute nodes by \emph{aprun}, Cray's replacement for mpirun.  by default, aprun is designed to start a single executable in single-program-multiple-data (SPMD) mode, not to start two different executables.  Aprun does support multiple-program-multiple-data (MPMD) mode, but it will not start two different executables on the same node (i.e. each program executes on non-overlapping subsets of nodes).  The solution is to have one rank on each node call \texttt{fork()} and then \texttt{execve()}.  Since MPI purposely hides the compute topology from the application, deciding exactly which ranks call fork() takes a little more work.  Specifically, all ranks will attempt to open a file using the \texttt{O\_CREAT} and 
\texttt{O\_EXCL} flags.  The name of the file is taken from the compute node's host name.  This combination of flags and filename ensures that exactly one rank on each node successfully opens the file.  The ranks that succeed will then start the daemon processes.

>>>>

\subsection{Overview}
\label{subsec:overview}
To test this idea, the authors created a synthetic benchmark application that consists of two executables: the main application which uses MPI and simulates the I/O patterns of a typical HPC application and a daemon process that is started on each compute node.  The daemon has two tasks: manage the GPU memory and copy data from GPU memory out to a file.  Note that the daemon does \emph{not} copy the data into the GPU memory; that is handled by the individual ranks of the main application. There needs to be exactly one daemon process on each compute node regardless of how many MPI ranks are running on each node.\footnote{Due to the limitations of aprun, starting this daemon is somewhat involved.  See section \ref{subsec:cray-specific}.}

@@

The daemon maintains a pair of lists.  One list is the blocks of GPU memory that have been assigned to the ranks of the main application, and which the ranks are presumably copying data into.  The second list is for all the blocks of memory for which the `write complete' message (step e) has been received.  When the daemon thread that handles CCI messages receives a `write done' message for a block, it moves that block from the first list to the second list.  It then notifies the background write thread(s) that the block is now ready to be written to the filesystem.

@@

There are a few peculiarities of the Cray environment that need to be taken in to account in order for this software to work properly.  The first peculiarity has to do with limiting access to the GPU and associated hardware.  The CUDA runtime can be configured to limit access to the GPU to a single thread, to multiple threads from a single process, or to multiple threads from multiple processes.  On Titan, if no special options are given to qsub, the CUDA runtime will be configured for a single thread.  Since software described in this paper requires access to the GPU from multiple processes, users must switch to the appropriate compute mode.  This is done by passing the flag ``-l feature=gpudefault'' to qsub.\footnote{It may seem strange to have to explicitly switch to a mode called `gpudefault'.  The naming scheme comes from the CUDA documentation.  On a standard CUDA install, the GPU would in fact default to allowing access from multiple processes.  Titan is configured differently, however.  Thus the need to explicitly tell qsub to switch the GPUs to the `default' mode.}

The second peculiarity has to do with actually starting the daemon process.  Normally, executables are started on the compute nodes by \emph{aprun}, Cray's replacement for mpirun.  By default, aprun is designed to start a single executable in single-program-multiple-data (SPMD) mode, not to start two different executables.  Aprun does support multiple-program-multiple-data (MPMD) mode, but it will not start two different executables on the same node (i.e. each program executes on non-overlapping subsets of nodes).  The solution is to have one rank on each node call \texttt{fork()} and then \texttt{execve()}.  Since MPI purposely hides the compute topology from the application, deciding exactly which ranks call fork() takes a little more work.  Specifically, all ranks will attempt to open a file using the \texttt{O\_CREAT} and 
\texttt{O\_EXCL} flags.  The name of the file is taken from the compute node's host name.  This combination of flags and filename ensures that exactly one rank on each node successfully opens the file.  The ranks that succeed will then start the daemon processes.

### 2015-04-07 cug_paper/03_analysis.tex
Paper: Minor edits

For the second test series, the application was configured for 16 ranks per node.  The daemon was again run with 1, 2, and 4 write threads.  For this series, the daemon thread that handled the CCI messages was run in a mode where it would block waiting on a message.  This added some latency to the message processing, but left that core free to perform useful work when there were no messages to process.  Also for this series, no core pinning was used on the daemon.\footnote{In practical terms, a user would probably not want to use multiple write threads on the daemon if his/her applications was running 16 ranks/node since it would oversubscribe the cores.  The authors tested the daemon with 2 and 4 threads partially out of curiosity and also to keep the two test series as similar to each other as possible.}

>>>>

For the second test series, the application was configured for 16 ranks per node.  The daemon was again run with 1, 2, and 4 write threads.  For this series, the daemon thread that handled the CCI messages was run in a mode where it would block waiting on a message.  This added some latency to the message processing, but left that core free to perform useful work when there were no messages to process.  Also for this series, no core pinning was used on the daemon.\footnote{In practical terms, a user would probably not want to use multiple write threads on the daemon if his/her application was running 16 ranks/node since it would oversubscribe the cores.  The authors tested the daemon with 2 and 4 threads partially out of curiosity and also to keep the two test series as similar to each other as possible.}
